---
title: "An introduction to biodbChebi"
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodbChebi')`"
abstract: |
  How to use the biodbChebi connector and its methods.
vignette: |
    %\VignetteIndexEntry{Introduction to the biodbChebi package.}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output:
    BiocStyle::html_document:
        toc: yes
        toc_depth: 4
        toc_float:
            collapsed: false
    BiocStyle::pdf_document: default
bibliography: references.bib
---

# Introduction

biodbChebi is a *biodb* extension package that implements a connector to
the [ChEBI](https://www.ebi.ac.uk/chebi/) database [@degtyarenko2007_ChEBI; @hastings2012_chebi].

# Installation

Install using Bioconductor:
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install('biodbChebi')
```

# Initialization

The first step in using *biodbChebi*, is to create an instance of the biodb
class `BiodbMain` from the main *biodb* package. This is done by calling the
constructor of the class:
```{r, results='hide'}
mybiodb <- biodb::newInst()
is_online <- tryCatch({
    url <- "https://www.ebi.ac.uk/webservices/chebi/2.0/webservice?wsdl"
    res <- RCurl::getURL(url, .opts = list(timeout = 5, connecttimeout = 5))
    grepl("^\\s*<\\?xml|<wsdl:definitions", res)
}, error = function(e) FALSE)
```
During this step the configuration is set up, the cache system is initialized
and extension packages are loaded.

We will see at the end of this vignette that the *biodb* instance needs to be
terminated with a call to the `terminate()` method.

# Creating a connector to ChEBI database

In *biodb* the connection to a database is handled by a connector instance that
you can get from the factory.
biodbChebi implements a connector to a local database, thus when creating an
instance you must provide a URL that points to your database:
```{r, eval=is_online}
chebi <- mybiodb$getFactory()$createConn('chebi')
```

# Requesting entries

Using accession numbers, we request entries from ChEBI:
```{r, eval=is_online}
entries <- chebi$getEntry(c('2528', '17053', '15440'))
```

# Get entry fields

Getting the values of entry fields are done using `getFieldValue()` method.
Here we retrieve the SMILE field of the first entry from the ChEBI entries
obtained previously:
```{r, eval=is_online}
entries[[1]]$getFieldValue('smiles')
```

# Get data frame

We can convert an entry into a data frame:
```{r, eval=is_online}
entries[[1]]$getFieldsAsDataframe()
```

Building a data frame for a list of entries is also possible:
```{r, eval=is_online}
mybiodb$entriesToDataframe(entries, fields=c('accession', 'formula',
    'molecular.mass', 'inchikey', 'kegg.compound.id'))
```

# Search by name and mass

Searching by name and/or mass is done through the `searchCompound()`.

Searching by name:
```{r, eval=is_online}
chebi$searchCompound(name='aspartic', max.results=3)
```

Searching by mass:
```{r, eval=is_online}
chebi$searchCompound(mass=133, mass.field='molecular.mass', mass.tol=0.3,
    max.results=3)
```

Searching by name and mass:
```{r, eval=is_online}
ids <- chebi$searchCompound(name='aspartic', mass=133,
    mass.field='molecular.mass', mass.tol=0.3, max.results=3)
```
Display results in data frame:
```{r, eval=is_online}
mybiodb$entriesToDataframe(chebi$getEntry(ids), fields=c('accession',
    'molecular.mass', 'name'))
```

# Convert CAS IDs

Converting CAS IDs to ChEBI IDs:
```{r, eval=is_online}
chebi$convCasToChebi(c('87605-72-9', '51-41-2'))
```

If more than one ChEBI ID is found for a CAS ID, then a list of character
vectors is returned:
```{r, eval=is_online}
chebi$convCasToChebi('14215-68-0')
```
This behaviour can be made the default one by setting `simplify` parameter to
`FALSE`.

# Convert InChI

The method is similar to `convCasToChebi()`.

Converting InChI to ChEBI IDs:
```{r, eval=is_online}
chebi$convInchiToChebi('InChI=1S/C8H11NO3/c9-4-8(12)5-1-2-6(10)7(11)3-5/h1-3,8,10-12H,4,9H2/t8-/m0/s1')
```

You can also use an InChI key:
```{r, eval=is_online}
chebi$convInchiToChebi('MBDOYVRWFFCFHM-SNAWJCMRSA-N')
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```

# Session information

```{r}
sessionInfo()
```

# References
